Discrete Quantum Theories and Computing

### 2018-03-12 dissertation.tex
dissertation.lyx "2.1 Geometrical Structure of States" Updating...

In summary, the full space of a $D$-dimensional quantum state, including
its overall phase defining its relationship to other quantum states,
is the topological space $\Sphere{2D-1}$. For an isolated system,
the overall phase is not measurable, and eliminating the phase dependence
in turn corresponds to identifying $\Sphere{2D-1}$ as a circle bundle
over the base space $\CP{D-1}$, and therefore $\CP{D-1}$ defines
the $2D-2$ intrinsic, irreducible, degrees of freedom of the isolated
$D$-dimensional state's dynamics. In mathematical notation, this
would be written $\Sphere{1}\hookrightarrow\Sphere{2D-1}\rightarrow\CP{D-1}$.
For $D=2$, the single qubit, we have $2-1=1$, and the base space
of the circle bundle is $\CP{1}=\Sphere{2}$, the usual Bloch sphere.
Note that only for $D=2$ is this actually a sphere-like geometry
due to an accident of low-dimensional topology.

@@

Given a normalized pure state $\ket{\Psi}=\sum_{i=0}^{D-1}\alpha_{i}\ket{i}$,
a natural quantity characterizing an $D$-dimensional system is its
\textit{density matrix}, $\rho=\op{\Psi}{\Psi}$, or 
\begin{equation}
\rho=\begin{bmatrix}|\alpha_{0}|^{2} & \alpha_{0}\alpha_{1}^{*} & \cdots & \alpha_{0}\alpha_{D-1}^{*}\\
\alpha_{1}\alpha_{0}^{*} & |\alpha_{1}|^{2} & \cdots & \alpha_{1}\alpha_{D-1}^{*}\\
\vdots & \vdots & \ddots & \vdots\\
\alpha_{D-1}\alpha_{0}^{*} & \cdots & \alpha_{D-1}\alpha_{D-2}^{*} & |\alpha_{D-1}|^{2}
\end{bmatrix}\,.\label{nqDensity.eq}
\end{equation}
We can now use the complex generalization of the classical Veronese
coordinate system for projective geometry to remove the overall phase
ambiguity $\rme^{\rmi\theta}$ from the $D$-dimensional states. If
we take a particular weighting of the elements of the density matrix~$\rho$,
we can construct a \textit{unit vector} of real dimension $D^{2}$
with the form: 
\begin{equation}
\Hat{a}=\left(|\alpha_{i}|^{2},\ldots,\sqrt{2}\,\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\,\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots\right)\,,\label{nqCoords.eq}
\end{equation}
where 
\begin{equation}
\Hat{a}\cdot\Hat{a}=\sum_{i=0}^{D-1}\left(\left|\alpha_{i}\right|^{2}\right)^{2}+\sum_{i=0}^{D-1}\sum_{\substack{j=0\\
j\ne i
}
}^{D-1}\left(\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*}\right)^{2}+\left(\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*}\right)^{2}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)\left(\sum_{j=0}^{D-1}\left|\alpha_{j}\right|^{2}\right)=1\,.
\end{equation}
This construction gives an explicit embedding of the $\left(D-1\right)$-dimensional
complex, or $\left(2D-2\right)$-dimensional real, object in a real
space of dimension $D^{2}$. However, this is somewhat subtle because
the vector is of unit length, so technically the embedding space is
a sphere of dimension $D^{2}-1$ embedded in $\R^{D^{2}}$. For example,
the two-dimensional irreducible states could be represented in a four-dimensional
embedding, but the magnitude of every coordinate would be one; furthermore,
the object embedded in the resulting $\Sphere{3}$ is indeed $\Sphere{2}$
because we can fix one complex coordinate to be unity, and let one
vary, giving a total of two irreducible dimensions. In fact one must
choose \textit{two} coordinate patches, one covering one pole of $\Sphere{2}$
with coordinates $\alpha_{0}=0$ and $\alpha_{1}=x_{1}+\rmi y_{1}$,
and the other patch covering the other pole of $\Sphere{2}$ with
coordinates $\alpha_{0}=x_{0}+\rmi y_{0}$ and $\alpha_{1}=0$.

@@

\todo{{Need to define }(partial) trace and mixed state before this
subsection...}

>>>>

In summary, the full space of a $D$-dimensional quantum state, including
its overall phase defining its relationship to other quantum states,
is the topological space $\Sphere{2D-1}$. For an isolated system,
the overall phase is not measurable, and eliminating the phase dependence
in turn corresponds to identifying $\Sphere{2D-1}$ as a circle bundle
over the base space $\CP{D-1}$, and therefore $\CP{D-1}$ defines
the $2D-2$ intrinsic, irreducible, degrees of freedom of the isolated
$D$-dimensional state's dynamics. In mathematical notation, this
would be written $\Sphere{1}\hookrightarrow\Sphere{2D-1}\rightarrow\CP{D-1}$\todo{Citation?}.
For $D=2$, the single qubit, we have $2-1=1$, and the base space
of the circle bundle is $\CP{1}=\Sphere{2}$, the usual Bloch sphere.
Note that only for $D=2$ is this actually a sphere-like geometry
due to an accident of low-dimensional topology.

@@

Given a normalized pure state $\ket{\Psi}=\sum_{i=0}^{D-1}\alpha_{i}\ket{i}$,
a natural quantity characterizing an $D$-dimensional system is its
\textit{density matrix}, $\rho=\op{\Psi}{\Psi}$, or 
\begin{equation}
\rho=\begin{pmatrix}|\alpha_{0}|^{2} & \alpha_{0}\alpha_{1}^{*} & \cdots & \alpha_{0}\alpha_{D-1}^{*}\\
\alpha_{1}\alpha_{0}^{*} & |\alpha_{1}|^{2} & \cdots & \alpha_{1}\alpha_{D-1}^{*}\\
\vdots & \vdots & \ddots & \vdots\\
\alpha_{D-1}\alpha_{0}^{*} & \cdots & \alpha_{D-1}\alpha_{D-2}^{*} & |\alpha_{D-1}|^{2}
\end{pmatrix}\,.\label{nqDensity.eq}
\end{equation}
We can now use the complex generalization of the classical Veronese
coordinate system for projective geometry to remove the overall phase
ambiguity $\rme^{\rmi\theta}$ from the $D$-dimensional states. If
we take a particular weighting of the elements of the density matrix~$\rho$,
we can construct a \textit{unit vector} of real dimension $D^{2}$
with the form: 
\begin{equation}
\Hat{a}=\left(|\alpha_{i}|^{2},\ldots,\sqrt{2}\,\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\,\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots\right)\,,\label{nqCoords.eq}
\end{equation}
where 
\begin{equation}
\Hat{a}\cdot\Hat{a}=\sum_{i=0}^{D-1}\left(\left|\alpha_{i}\right|^{2}\right)^{2}+\sum_{i=0}^{D-1}\sum_{\substack{j=0\\
j\ne i
}
}^{D-1}\left(\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*}\right)^{2}+\left(\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*}\right)^{2}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)\left(\sum_{j=0}^{D-1}\left|\alpha_{j}\right|^{2}\right)=1\,.
\end{equation}
This construction gives an explicit embedding of the $\left(D-1\right)$-dimensional
complex, or $\left(2D-2\right)$-dimensional real, object in a real
space of dimension $D^{2}$. However, this is somewhat subtle because
the vector is of unit length, so technically the embedding space is
a sphere of dimension $D^{2}-1$ embedded in $\R^{D^{2}}$. For example,
the two-dimensional irreducible states could be represented in a four-dimensional
embedding, but the magnitude of every coordinate would be one; furthermore,
the object embedded in the resulting $\Sphere{3}$ is indeed $\Sphere{2}$
because we can fix one complex coordinate to be unity, and let one
vary, giving a total of two irreducible dimensions. In fact one must
choose \textit{two} coordinate patches, one covering one pole of $\Sphere{2}$
with coordinates $\alpha_{0}=1+\rmi0$ and $\alpha_{1}=x_{1}+\rmi y_{1}$,
and the other patch covering the other pole of $\Sphere{2}$ with
coordinates $\alpha_{0}=x_{0}+\rmi y_{0}$ and $\alpha_{1}=1+\rmi0$.
\todo{Explain the last part more clear or add a picture.}

@@

\todo{Need to define (partial) trace and mixed state before this
subsection...}
